\documentclass[fleqn,10pt,lineno]{wlpeerj}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{placeins}
\usepackage{float}
\usepackage{flafter}
\usepackage{microtype}
\usepackage{url}
\usepackage[authoryear,round]{natbib}
\usepackage{hyperref}
\usepackage{graphicx}

\let\cite\citep

\graphicspath{{../}{figures/}{../figures/}}

\title{Matrix-Residual Graph Neural Networks for Graph Classification}

\author[1]{Xuelin Hu}
\author[2]{Xiaoqin Fu}

\affil[1]{School of Railway Communication and Signaling Engineering, Liuzhou Railway Vocational Technical College, Liuzhou 545000, China}
\affil[2]{School of Railway Locomotive and Rolling Stock, Liuzhou Railway Vocational Technical College, Liuzhou 545000, China}

\corrauthor[2]{Xiaoqin Fu}{fuxiaoqin@ltzy.edu.cn}

\begin{abstract}
\textbf{Background.} Residual connections are widely used to stabilize graph neural networks, but residual design is often treated as a one-dimensional depth-wise choice. Multi-branch graph classifiers create a richer design space: residual information can flow along depth, across branches, or through a local two-dimensional branch-by-layer neighborhood.

\textbf{Methods.} We study Matrix-Residual Graph Neural Networks as a controlled residual-topology framework for graph classification. The model family represents hidden states on a branch-by-layer grid and compares five information-flow designs: plain message passing, vertical residual reuse, horizontal branch-wise exchange, local matrix residual reuse, and sparse/gated matrix reuse. The main benchmark evaluates six datasets, four message-passing operators, five model families, and five graph-level folds under a shared training protocol. We further run branch-count ablations, fold-0 sensitivity scans, five-fold checks of selected MatrixResGated candidates, and mechanism analyses based on branch diversity, branch cosine similarity, branch CKA, gradient norms, residual activity, and learned gates.

\textbf{Results.} No single residual topology dominates all datasets and operators. Across the 24 dataset--operator combinations, MatrixRes obtains the largest number of wins (9/24) and the best mean rank, followed by MatrixResGated. Under GCNConv, the winners differ by dataset: Plain is strongest on PROTEINS, HorizontalRes is strongest on DD, MatrixResGated is strongest on the low-accuracy ENZYMES setting, and MatrixRes is strongest on MUTAG, AIDS, and Mutagenicity. Branch-count ablations on PROTEINS and DD reveal dataset-dependent operating regions rather than monotonic capacity gains. Mechanism metrics are consistent with the view that useful branch expansion needs representational separation without excessive residual traffic or weakened gradients. Five-fold tuned-candidate checks show that a smaller MatrixResGated hidden dimension reduces parameters on DD while retaining competitive accuracy; the PROTEINS sparsity candidate remains variable across folds.

\textbf{Conclusions.} Residual topology, branch count, and residual filtering should be selected jointly. Additional branches help only when they create useful functional diversity; otherwise, residual traffic and branch redundancy can offset the added capacity. Matrix-style reuse is best understood as a tunable residual family rather than a universal replacement for simpler vertical or horizontal residual paths.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\input{sections/01_introduction_peerj.tex}
\input{sections/02_methods_peerj.tex}
\input{sections/03_results_peerj.tex}
\input{sections/04_discussion_peerj.tex}
\input{sections/05_conclusions_peerj.tex}
\input{sections/06_acknowledgments.tex}
\input{sections/07_appendix.tex}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
